Application of Turchin's method of statistical regularization

### 2018-03-08 AYSS-2017/article/ayss2017-Zeleniy.tex

We expect that $\varphi$ is relatively smooth and doesn't contain wild
oscillations which naive least squares usually produce. Common way of measuring
function's smoothness is $\langle\varphi,\hat{\Omega}\varphi\rangle$, where
$\hat{\Omega}=|\frac{d^2}{dx^2}\left\rangle\right\langle\frac{d^2}{dx^2}|$.
At the same time we want to minimize information content of prior.

Reasonable choice is ensemble of functions with average smoothness $\omega$ which
maximizes Shannon entropy. Which means we need to maximize:

@@

The resulting prior probability density for the Gaussian random process:
\begin{equation}
  P(\vec{\varphi}|\alpha) = \frac{\det(\alpha\Omega)^{1/2}}{(2\pi)^{N/2}} 
  \exp\left(-\frac{1}{2}\langle\vec{\varphi},\alpha\Omega\vec{\varphi}\rangle\right),
  \label{eq-prior}
\end{equation}

where $\alpha = \alpha(\omega)$~--- parameter of smoothness. Now we have problem
of picking $\alpha$. Information about its value almost never available so we
could consider it random variable as well. This yields us hierarchical Bayes
model:

@@

\begin{enumerate}
\item Use fully Bayesian approach and marginalize $\alpha$:
  \begin{equation}\label{eq:varAposteriorAlpha}
      P(\varphi) = \int P(\varphi|\alpha) P(\alpha)~d\alpha
  \end{equation}
  In general case MCMC could be used to sample $\varphi$ from full
  posterior. And for Gaussian errors it could be reduced to numerical integral.

\item Use empirical Bayes: pick $\alpha$ corresponding to most probable
  $P(\alpha|f)$

  \begin{equation}
    P(\alpha|f) = \int d\varphi~P(f|\varphi)P(\varphi|\alpha)P(\alpha)
  \end{equation}

  If distribution $P(\alpha|f)$ is narrow it will produce approximately same
  result. And it's the case for our example problem
\end{enumerate}

@@

In order to find numerical solution, one need to go from continuous functions in eq.~\ref{eq-02} to discrete matrix form:
\begin{equation}\label{eq:algebr}
	\begin{gathered}
      \varphi(x) = \sum \limits_n \varphi_n T_n(x),\\
      K_{mn} = (\hat{K}T_n(x))(y_m),\\
      \Omega_{ij} = \int\limits_a^b \left(\frac{d^pT_i(x)}{dx}\right)\left(\frac{d^pT_j(x)}{dx}\right)dx,\\
      D[\varphi(x)] = D[\sum \limits_n \varphi_n T_n(x)] = \sum\limits_{i,j} \varphi_i\varphi_j cov(T_i(x), T_j(x)),\\
      \vec{f} = K\vec{\varphi},
    \end{gathered}
\end{equation}
where $T_n(x)$ - some basis in function space. For example, cubic spline, Fourier series and Legendre polynomials.

>>>>

We expect that $\varphi$ is relatively smooth and doesn't contain wild
oscillations which, for example, we usually produce in use naive least squares. Common way of measuring
function's smoothness is $\langle\varphi,\hat{\Omega}\varphi\rangle$, where
$\hat{\Omega}=|\frac{d^2}{dx^2}\left\rangle\right\langle\frac{d^2}{dx^2}|$.
At the same time we want to minimize information content of prior.

Reasonable choice is ensemble of functions with average smoothness $\omega$ which
maximizes Shannon entropy (or is equivalent to minimize that value of Shannon's information, which we add to solution). Which means we need to maximize:

@@

The resulting prior probability density for the Gaussian random process:
\begin{equation}
  P(\vec{\varphi}|\alpha) = \frac{\det(\alpha\Omega)^{1/2}}{(2\pi)^{N/2}} 
  \exp\left(-\frac{1}{2}\langle\vec{\varphi},\alpha\Omega\vec{\varphi}\rangle\right),
  \label{eq-prior}
\end{equation}
where $\alpha = \alpha(\omega)$~--- parameter of smoothness. In result, we get $\alpha$-depend solution $hat{S}_{\alpha}[f]$ and now we have problem of picking $\alpha$. In rare cases we can select $alpha$ manually using known empirical smoothness, but we proposal another way. Information about its value almost never available so we
could consider it random variable as well. This yields us hierarchical Bayes
model:

@@

\begin{enumerate}
    \item Use fully Bayesian approach and marginalize $\alpha$:
    \begin{equation}\label{eq:varAposteriorAlpha}
    P(\varphi) = \int P(\varphi|\alpha) P(\alpha)~d\alpha
    \end{equation}
    \item Use empirical Bayes: pick $\alpha$ corresponding to most probable
    $P(\alpha|f)$:
    \begin{equation}
    P(\alpha|f) = \int d\varphi~P(f|\varphi)P(\varphi|\alpha)P(\alpha)
    \end{equation}  
\end{enumerate}
In general case this options is equivalent: best solution definition by next formula:
\begin{equation}
\hat{S}[f] = \frac{1}{Norm} \int d\alpha \hat{S}_{\alpha}[f] P(\alpha)~d\alpha =\frac{1}{Norm} \int d\alpha \hat{S}_{\alpha}[f] P(\alpha|f)
\end{equation}
and MCMC could be used to sample $\varphi$ and $alpha$ for computing solution (for Gaussian errors it could be reduced to numerical integral). In private case this options can give different solutions if we will use some approximations. For example we can used most probability $alpha$ by $P(\alpha|f)$. This method used for our application problem.

@@

In order to find numerical solution, one need to go from continuous functions in eq.~\ref{eq-02} to discrete matrix form:
\begin{equation}\label{eq:algebr}
	\begin{gathered}
      \varphi(x) = \sum \limits_n \varphi_n T_n(x),
      \qquad\mbox{and}\qquad
      K_{mn} = (\hat{K}T_n(x))(y_m),\\
      \Omega_{ij} = \int\limits_a^b \left(\frac{d^pT_i(x)}{dx}\right)\left(\frac{d^pT_j(x)}{dx}\right)dx,\\
      D[\varphi(x)] = D[\sum \limits_n \varphi_n T_n(x)] = \sum\limits_{i,j} \varphi_i\varphi_j cov(T_i(x), T_j(x)),\\
      \vec{f} = K\vec{\varphi},
    \end{gathered}
\end{equation}
where $T_n(x)$ - some basis in function space. For example, cubic spline, Fourier series and Legendre polynomials.
